---
title: "An Exploration of the Central Limit Theorem Relating to Sample Mean and Sample 
Variance of Random Variables from R Language's Pseudorandom Number Generator and its
Implementation of Exponential Distribution"
author: "Randall"
output: pdf_document
---

```{r global_options, echo = FALSE, message}
library(knitr)
opts_chunk$set(cache = TRUE)
library(dplyr) # loads magrittr and forward pipe operator `%>%` 
library(xtable) # better knitr table printing
```

# Theoretical and Sample Mean of the Exponential Distribution

The exponential distribution describes the elapsed time between events 
in a Poisson process. Poisson processes are characterized by "memorylessness", 
which means that while the probability of total events having occurred accrues
over time, the probability of an event occurring within any given time period 
is not affected by occurrences within other periods of time.

Where $x$ is the elapsed time in units, $\lambda$ is the rate, in events per unit of
time. The probability density function of the exponential distribution which describes
the elapsed time between events for $x \ge 0$ is $F(x;\lambda) = \lambda e^{-\lambda x}$
and its CDF is $F(x;\lambda) = 1-e^{-\lambda x}$.

By definition, $1/\lambda$ corresponds to the expected time between events. For example, 
if 3600 hits are expected to a website every hour, then it is also expected that the 
rate will *average* to one hit per second.

>Therefore:  
>If $\lambda = 0.2 = 1/5$  
>then $E[X] = 1/\lambda = 5$, the center of the distribution of elapsed times

```{r}
set.seed(123); lambda <- 0.2; num <- 40
exp.sample40 <- rexp(n = num, rate = lambda)
for (i in 1:999) {
    exp.sample40 <- cbind(exp.sample40,
                          rexp(n = num, rate = lambda)
                          )
}

```

