---
title: "An Exploration of Tooth Growth"
author: "Randall"
date: "January 25, 2015"
output: pdf_document
---

```{r global_options, echo = FALSE, message = FALSE}
library(knitr)
opts_chunk$set(cache = TRUE)
```

The data for this exploration is documented:

>The response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).

The variables are `len`, tooth length in mm, `supp`, whether vitamin C was administered
as orange juice (OJ) or as ascorbic acid (VC), and `dose` in mg of vitamin C.

For the purposes of this project I will treat tooth length as a dependent variable of
dose and supplement type.

```{r setup, message=FALSE}
library("datasets"); library("ggplot2"); library("UsingR")
library("tidyr") # change shape of data
library("dplyr") # data manipulation and loads magrittr
library("magrittr")
library("xtable") # formats tables in knitr
data("ToothGrowth")
```
```{r ToothGrowth exploration, results='asis'}
ToothGrowth %<>% tbl_df # convert ToothGrowth promise into dplyr::tbl_df object
ToothGrowth[sample(x = 60, size = 5), ] %>% 
    xtable %>% print(comment=FALSE) #Show random 5 values
```

A quick exploratory plot of the data give an indication what is good to study further:
```{r Exploratory Plot, warning=FALSE, message=FALSE, fig.height=2}
qplot(data = ToothGrowth, x = dose, y = len, color = supp) + geom_smooth()
```

Notably, it looks like increased Vitamin C is associated with longer tooth length, and
at least at lower doses, that orange juice is a more effective delivery method than
ascorbic acid by itself.

# Association Between Vitamin C and Tooth Length

Let's hypothesize[^1] that a dose of 1mg of vitamin C is associated with longer teeth in 
guinea pigs than a vitmain C dose of 0.5mg.

* $H_{0} : \mu_{length, dose=1mg} \leq \mu_{length, dose = 0.5mg}$
* $H_{a} : \mu_{length, dose=1mg} > \mu_{length, dose = 0.5mg}$

```{r}
ToothGrowth %<>% group_by(dose)
len_dose_0.5 <- ToothGrowth %>% filter(dose == 0.5)
len_dose_1 <- ToothGrowth %>% filter(dose == 1)

t.test(len_dose_1$len - len_dose_0.5$len)
```

Yes, it is very safe to reject the null hypothesis that 1mg of vitamin C is not 
associated with longer tooth length than 0.5mg.

The 95% confidence interval says that the increased dosage of vitamin C corresponds 
with between 6.387 and 11.873 mm longer teeth. Furthermore the null hypothesis is 
rejected with a lot of power, it is rejected even with a p-value as low as 0.000001225.

Note that this assumes some additional properties. For instance that both populations
have similar variance:
```{r}
var(len_dose_1$len)
var(len_dose_0.5$len)
```

It also assumes that doses were chosen randomly, without any confounding reasons. For
example, if guinea pigs with longer teeth were larger at the beginning and had a larger 
appetite, they might consume more of the supplements. Though we should investigate the 
methodology, it seems unlikely that the researchers make this type of absurd mistake.

[^1]: This is not a legitimate way to conduct science. A hypothesis should be 
chosen before an experiment is performed, and then the experiment is used to find a 
weight of evidence. Choosing a hypothesis to match the data is *not* science, but it 
is an okay way to practice statistics for the sake of an assignment.